A Lock-Free Interaction Combinator Evaluator

### 2023-09-08 paper/draft.tex
fix bibliography
paper formatting improvements

\lstset{
    language=Python, 
    basicstyle=\ttfamily\small,
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2,
    frame=single,
    backgroundcolor=\color{backcolour}
}

@@

\begin{abstract}
We present a lock-free evaluator for interaction combinators, a simple yet powerful model of distributed computation. The proposed algorithm is based on the concept of implicit ownership regions and redirectors, allowing for efficient parallel evaluation without race conditions or deadlocks. We provide detailed pseudocode for the core procedures and illustrations. We have implemented our lock-free evaluator in CUDA and achieved successful reductions of large graphs on GPUs with thousands of concurrent threads, demonstrating the practicality and efficiency of our approach.
\end{abstract}

@@

\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{interaction_rules.png}

@@

\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_based_evaluator_1.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_based_evaluator_2.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_based_evaluator_3.png}

@@

Below is a complete illustration of this process:

\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_1.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_2.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_3.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_4.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_5.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_6.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_7.png}

Note that in Image 2, the reduction is semantically complete, with ports X1-W0 and Y0-Z0 indirectly connected. Each thread independently performs this reduction, without impacting surrounding nodes, by converting the auxiliary ports of their owned nodes to redirectors and then swapping opposing ports (e.g., A1 goes to B1, B2 goes to A2, etc.). This approach entirely avoids locks. The link() procedure is then only needed to clear the memory occupied by redirectors and to identify new active pairs. As this clean-up process may cross ownership regions, caution is required to deploy atomics when necessary. The following is a comprehensive pseudocode:

@@

- Move A1 as redirector to B1; if target is main, link B1 towards A1.

- Move A2 as redirector to B2; if target is main, link B2 towards A2.

- Move B1 as redirector to A1; if target is main, link A1 towards B1.

- Move B2 as redirector to A2; if target is main, link A2 towards B2.

These 4 segments are, themselves, independent, and can be performed in parallel. This allows a single interaction to be processed by 4 threads, each one performing 1/4 of a rewrite. According to Amdahl's law, the maximum speedup achievable through parallelization is limited by the fraction of the program that must be executed sequentially. By dividing the rewrite process into 4 smaller sequential chunks, we effectively reduce the sequential portion of the algorithm, increasing the maximum potential speedup by 4x. The code below demonstrates the implementation of this optimized approach, completing the algorithm and allowing for efficient parallel execution of interaction combinators.

@@

\bibliographystyle{plain}
\bibliography{references}

\bibitem{taelin2022lock} 
Taelin, V. (2022). Higher-order Virtual Machine [Online]. Available at: https://github.com/HigherOrderCO/hvm-core

>>>>

\lstset{
    language=Python, 
    basicstyle=\ttfamily\small,
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2,
    frame=single,
    backgroundcolor=\color{white}
}

@@

\begin{abstract}
We present a lock-free evaluator for interaction combinators, a simple yet powerful model of distributed computation. The proposed algorithm is based on the concept of implicit ownership regions and redirectors, allowing for efficient parallel evaluation without race conditions or deadlocks. We provide detailed illustrations, and the pseudocode for the core procedures can be found in the appendix. We have implemented our lock-free evaluator in CUDA and achieved successful reductions of large graphs on GPUs with thousands of concurrent threads, demonstrating the practicality and efficiency of our approach.
\end{abstract}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{interaction_rules.png}
\caption{Interaction rules}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_based_evaluator_1.png}
\caption{Lock-based evaluator step 1}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_based_evaluator_2.png}
\caption{Lock-based evaluator step 2}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_based_evaluator_3.png}
\caption{Lock-based evaluator step 3}
\end{figure}

@@

The following illustrations showcase this process:

\begin{figure}[h!]
\centering
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_1.png}
\caption{Lock-free evaluator step 1}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_2.png}
\caption{Lock-free evaluator step 2}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_3.png}
\caption{Lock-free evaluator step 3}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_4.png}
\caption{Lock-free evaluator step 4}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_5.png}
\caption{Lock-free evaluator step 5}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_6.png}
\caption{Lock-free evaluator step 6}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_7.png}
\caption{Lock-free evaluator step 7}
\end{figure}

Note that in Image 2, the reduction is semantically complete, with ports X1-W0 and Y0-Z0 indirectly connected. Each thread independently performs this reduction, without impacting surrounding nodes, by converting the auxiliary ports of their owned nodes to redirectors and then swapping opposing ports (e.g., A1 goes to B1, B2 goes to A2, etc.). This approach entirely avoids locks. The link() procedure in the appendix is then only needed to clear the memory occupied by redirectors and to identify new active pairs. As this clean-up process may cross ownership regions, caution is required to deploy atomics when necessary.

\begin{itemize}
  \item Move A1 as redirector to B1; if target is main, link B1 towards A1.
  \item Move A2 as redirector to B2; if target is main, link B2 towards A2.
  \item Move B1 as redirector to A1; if target is main, link A1 towards B1.
  \item Move B2 as redirector to A2; if target is main, link A2 towards B2.
\end{itemize}

These 4 segments are, themselves, independent, and can be performed in parallel. This allows a single interaction to be processed by 4 threads, each one performing 1/4 of a rewrite. According to Amdahl's law, the maximum speedup achievable through parallelization is limited by the fraction of the program that must be executed sequentially. By dividing the rewrite process into 4 smaller sequential chunks, we effectively reduce the sequential portion of the algorithm, increasing the maximum potential speedup by 4x. The pseudocode for this optimized approach can be found in the appendix, completing the algorithm and allowing for efficient parallel execution of interaction combinators.

\bibitem{taelin2022lock} 
Taelin, V. (2022). Higher-order Virtual Machine [Online]. Available at: https://github.com/HigherOrderCO/hvm-core
\end{thebibliography}

\appendix
\section{Pseudocode}\label{app:pseudocode}

In this appendix, we provide the complete pseudocode for the lock-free evaluator algorithm and the 1/4 rewrites optimization.

@@

@@
